---
title: "POA_accidents"
author: "arthur"
date: "October 31, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we need to download the dataset that contains the data about the accidents in Porto Alegre in 2016.

```{r}
file <- "acidentes-2016.csv"
if(!file.exists(file)){
  download.file("http://datapoa.com.br/storage/f/2017-08-03T13%3A19%3A45.538Z/acidentes-2016.csv",
	destfile=file)
}
```
  
Read the csv and put it in a dataframe
  
```{r}
library(readr);
df <- read_delim(file, delim=";");
head(df);
```


Now we will try to answer the questions:

  - Is there a time of the year with more accidents?
  - How many vehicles are usually involved?
  - Is there a specific weekday with more accidents?
  - Advanced: are there some area in POA with more accidents?
      - An area can be a neighborhood, an intersection. Define precisely what is your region in your Rmd.

To answer the first question, we will just count the entries by date.

```{r}
library(tidyverse)
library(ggplot2)

df %>%  group_by(DATA) %>% summarise(N=n()) %>% ggplot(aes(x=DATA, y=N)) + geom_line() + ylim(0,NA)
```

Visualizing by day is messy, let's try by month.

```{r}
df %>%  group_by(MES) %>% summarise(N=n())  %>% ggplot(aes(x=as.factor(MES), y=N)) + geom_point() + ylim(0,NA) + theme_bw() 
```

Actually, some months have more days than others, we must normalize the amount of crashes by the number of days in each month.

```{r}
library(lubridate)
df %>% group_by(MES) %>% summarize(normalized=n()/days_in_month(DATA) %>% head(1))  %>% ggplot(aes(x=as.factor(MES), y=normalized)) + geom_point() + ylim(0,NA) + theme_bw()
```

I can't believe that this acutally worked, let's verify:
```{r}
df %>% filter(MES==1) %>% summarise(N=n()) / 31
df %>% filter(MES==2) %>% summarise(N=n()) / 29
df %>% filter(MES==3) %>% summarise(N=n()) / 31

```

Seems legit :)

We can see that february is the month with the least crashes. One hypothesis is that it is the month when Porto Alegre is more desert because a lot of people go to other cities to enjoy the carnival.
March is the month with the most crashes.


It could be interesting to analyze by day of the week.

```{r}
dias_sem=c("DOMINGO", "SEGUNDA-FEIRA", "TERCA-FEIRA", "QUARTA-FEIRA", "QUINTA-FEIRA", "SEXTA-FEIRA", "SABADO")
df %>% group_by(DIA_SEM) %>% summarize(N=n()) %>% ggplot(aes(x=DIA_SEM, y=N)) + geom_point(stat = "identity") + ylim(0,NA) + scale_x_discrete(limits=dias_sem) + theme_bw()
```
We see that in the weekends there are not as many crashes as in the rest of the week. The days with the most crashes are mondays and fridays.

It could be intereseting to analyze by time of the day. 

Let's see the number of accidents, by weekday, from midnight to 6 AM. 

```{r}
df %>% filter(FX_HORA < 6) %>% group_by(DIA_SEM) %>% summarize(N=n()) %>% ggplot(aes(x=DIA_SEM, y=N)) + geom_point(stat = "identity") + ylim(0,NA) + scale_x_discrete(limits=dias_sem) + theme_bw()
```

Maybe the best way to represent this is to have a stacked bar plot with the days of the week on the x axis and the amount of accidents by period of the day like late night, day, night, etc. stacked forming the total amount of crashes.

First we create a dataframe with a variable that is the time of the day:

```{r}
df %>% filter(FX_HORA < 6) %>% group_by(DIA_SEM) %>% summarize(N=n()) -> late_night
late_night$time = "late_night"
df %>% filter(FX_HORA > 6 & FX_HORA < 19) %>% group_by(DIA_SEM) %>% summarize(N=n()) -> day
day$time = "day"
df %>% filter(FX_HORA > 19) %>% group_by(DIA_SEM) %>% summarize(N=n()) -> night
night$time = "night"
by_hour = rbind(day, rbind(late_night, night))
by_hour
```

```{r}
by_hour %>% ggplot() + geom_bar(aes(y=N, x=DIA_SEM, fill=time), stat="identity") + theme_bw() + scale_x_discrete(limits=dias_sem)
```
That does not look quite as good as I thought...

Let's try one plot of accidents by hour of the day

```{r}
df %>% group_by(FX_HORA) %>% summarise(N=n()) %>% ggplot(aes(x=as.factor(FX_HORA), y=N)) + geom_point() + theme_bw()
```

There is some problem with the data, maybe some observations do not have any value on FX_HORA

```{r}
df %>% select(FX_HORA) %>% summary()
```

There are 3 NA's, we can just filter them out and redo the plot

```{r}
df %>% group_by(FX_HORA) %>% filter(!is.na(FX_HORA)) %>% summarise(N=n()) %>% ggplot(aes(x=as.factor(FX_HORA), y=N)) + geom_point() + theme_bw()
```

That looks good. There are way more accidents between 8 AM and 17 PM than in other periods of the day. I would have guessed that 18 to 19 PM would be way worse.

It's enough for now, let's try to answer the other questions.

How many vehicles are usually involved?

```{r}
names(df)
```

Maybe CONT_ACID is the amount of vehicles

```{r}
df %>% select(CONT_ACID, AUTO, TAXI, LOTACAO, ONIBUS_URB, ONIBUS_MET, ONIBUS_INT, CAMINHAO, MOTO, CARROCA, BICICLETA, OUTRO) %>% head(3)
```

It's not. I must add all the vehicles.

```{r}
df %>%  group_by(ID) %>% summarize(vehicles = sum(AUTO, TAXI, LOTACAO, ONIBUS_URB, ONIBUS_MET, ONIBUS_INT, CAMINHAO, MOTO, CARROCA, BICICLETA, OUTRO)) %>% select(vehicles) -> by_vehicles 
by_vehicles %>% summary()
```
There is an accident with 0 vehicles?
```{r}
by_vehicles %>% filter(vehicles > 0 ) -> by_vehicles;
```

Quick plot to help us visualize this:
```{r}
by_vehicles %>% group_by(vehicles) %>% summarize(N=n()) %>% ggplot(aes(x=vehicles, y=N)) + theme_bw(base_size=22) + geom_bar(stat="identity")
```

Clearly most accidents happen between two vehicles.


About the locations, how can we define a location if all we have is coordinates and street names? Maybe let's try by street/avenue first:

The accidents on crossroads will count for both streets.

```{r}
df %>% filter(!is.na(LOG1)) %>% group_by(LOG1) %>% summarize(N=n()) -> log_1
df %>% filter(!is.na(LOG2)) %>% group_by(LOG2) %>% summarize(N=n()) -> log_2

# Sorry, I'm sure there is a much better way of doing this but this works for now and I would rather save time
names(log_2)[1] <- "LOG1"
rbind(log_1, log_2) %>% group_by(LOG1) %>% summarize(acc=sum(N)) -> acc_by_street

acc_by_street %>% arrange(-acc)

```

Good, we have the avenues with the most crashes in 2016. Maybe viewing the crossroads with the most accidents could be interesting?

```{r}
df %>% filter(!is.na(LOG2)) %>%  group_by(LOG1, LOG2) %>% summarize(N=n()) %>% arrange(-N)
```

Ipiranga has a lot of accidents but it also is huge so it's normal.

Maybe we can use the coordinates to view something more interesting. First we need to verify the integrity of the data.

```{r}
df %>% select(LONGITUDE, LATITUDE) %>% summary()
```

That huge value is obviously missing a dot. We could fix it but we could also just don't consider it.

```{r}
df %>% filter(LATITUDE > -60) %>% select(LATITUDE, LONGITUDE) %>% summary()
```

We still have -51.23 for latitude and -30.05 for longitude. Someone obviously put it in the wrong field, we must remove those. Let's define that all values for latitude must be between -29.5 and -30.5, and longitude between -50.8 and -51.8.

```{r}
df %>% filter(LATITUDE > -30.5 & LATITUDE < -29.5 & LONGITUDE > -51.8 & LONGITUDE < -50.8) %>% select(LATITUDE, LONGITUDE) -> df_filtered
df_filtered %>% summary
```
Since the coordinates are heavily rounded and there is a maximum of 25 values for latitude and 21 for longitude, maybe if I group the accidents by latitude and longitude I can know the "squares" with the most accidents 

```{r}
df_filtered %>% filter(LATITUDE > -30.5 & LATITUDE < -29.5 & LONGITUDE > -51.8 & LONGITUDE < -50.8) %>% group_by(LATITUDE, LONGITUDE) %>% summarize(N=n()) %>% arrange(-N)
```

Well, it's the summary function that rounds it... I think I will have to round it by hand. 

```{r}
df_rounded = df_filtered
df_rounded$LATITUDE = round(df_filtered$LATITUDE, 3)
df_rounded$LONGITUDE = round(df_filtered$LONGITUDE, 3)
df_rounded %>% group_by(LATITUDE,LONGITUDE) %>% summarise(N=n()) -> df_rounded
df_rounded %>% arrange(-N)

```

Good. Now I want to plot this as a heatmap.

```{r}
df_rounded %>% ggplot(aes(x=LONGITUDE, y=LATITUDE)) + geom_tile(aes(fill=N), colour="white") + scale_fill_gradient(low = "green",    high = "red") + theme_bw()
```

Wow... much better than expected. This acutally looks like Porto Alegre. We just have to fix the proportions.

Latitude has an amplitude of about 25 and latitude 21, so an aspect ratio of 25/21? Let's try.

```{r}
df_rounded %>% ggplot(aes(x=LONGITUDE, y=LATITUDE)) + geom_tile(aes(fill=N), colour="white") + scale_fill_gradient(low = "green",    high = "red") + theme_bw() + coord_fixed(25/21)
```

That looks better. Now we can zoom into the center. Let's see where are the places with more than 25 accidents.

```{r}
df_rounded %>% filter(N > 25) %>% summary()
```

it's between -30.10 and -30.00 latitude and -51.23 and -51.15 longitude. 
Also that gives us an aspect ratio of  10/8 or 5/4

```{r}
df_rounded %>% ggplot(aes(x=LONGITUDE, y=LATITUDE)) + geom_tile(aes(fill=N), colour="white") + scale_fill_gradient(low = "green",    high = "red") + theme_bw() + coord_cartesian(ylim = c(-30.10, -30.00), xlim = c(-51.23, -51.15)) 
```

That looks bad and "+ coord_fixed(5/4)" messes with it, I think the previous plot is already good enough.




